XY: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-03-27 paper/bare_conf.tex
finished writing licm impact

\subsection{Experimental Setup}
Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi. The \texttt{icc 2013} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$. Burgers is a time-dependent problem, i.e. the assembly is recalculated every time step based on the result of previous iterations. It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

@@

\begin{table*}[t]\normalsize
\massmatrixtwodresultsnorms
\label{table:perf-results-licm-helmholtz}
\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$. Each entry indicates the range of speed-ups obtained over the non-optimized implementation. The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding. Results are shown for both the Sandy Bridge and the Phi machine.}
\end{table*}

Tables~\ref{table:perf-results-licm-helmholtz},~\ref{table:perf-results-licm-advdiff}, and~\ref{table:perf-results-licm-burgers} illustrate the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code. Inspection of assembly code generated by $icc$ confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized. This motivates the usually significant gain. Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization. As shown in Figure~\ref{algo:PyOP2Compiler}, these transformations are applied on top of $licm$, so we refer to them as $licm$-$ap$. Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned. Occasionally $licm$-$ap$ is slower than $licm$ (e.g. in Burgers 3D p3). This is due to the large number of aligned temporaries introduced by $licm$, which probably leads to cache associativity conflicts.

\subsection{Impact of Vector-register Tiling}
Figures~\ref{fig:opvect-helmholtz-speedup},~\ref{fig:opvect-diffusion-speedup}, and~\ref{fig:opvect-burgers-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz, Diffusion, and Burgers assembly kernels, respectively. For each problem instance we report the best run-time obtained by empirically testing a set of different unroll/unroll-and-jam factors, and the one retrieved through the XY's cost model.

@@

\subsection{Impact of Expression Splitting}
% why split useless in helmholtz/diffusion
% why vect vect+split useless in burgers

\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
\label{fig:opvect-helmholtz-speedup}}
\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmholtz kernel.}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
\label{fig:opvect-diffusion-speedup}}
\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Diffusion kernel.}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
\label{fig:opvect-burgers-speedup}}
\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Burgers kernel.}
\end{figure}

>>>>

\subsection{Experimental Setup}
Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi. The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$. Burgers is a time-dependent problem, i.e. the assembly is recalculated every time step based on the result of previous iterations. It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

@@

\begin{table*}[t]\normalsize
\massmatrixtwodresultsnorms
\label{table:perf-results-licm}
\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$. Each entry indicates the range of speed-ups obtained over the non-optimized implementation. The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding. Results are shown for both the Sandy Bridge and the Phi machine.}
\end{table*}

Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code. Inspection of assembly code generated by $icc$ confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized. This motivates the usually significant gain obtained. Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization. As shown in Figure~\ref{algo:PyOP2Compiler}, these transformations are applied on top of $licm$, so we refer to them as $licm$-$ap$. Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned. Occasionally $licm$-$ap$ is slower than $licm$ (e.g. in Burgers 3D p3). This is due to the large number of aligned temporaries introduced by $licm$, which probably leads to cache associativity conflicts.

\subsection{Impact of Vector-register Tiling}
Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively. For each problem instance we report two bars: one indicates the best run-time achieved by autotuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the XY's cost model. In general, there is no substantial difference between the two. This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple. The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful. This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short. On the other hand, as soon as the iteration space becomes too big, vector-register tiles are essential to minimize pipeline stalls. A maximum speed-up of 1.40$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$. Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of arrays that, generally, need to be tiled. Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.

@@

\subsection{Impact of Expression Splitting}
\label{sec:perf-results-split}
Expression splitting, or loop fission, contributes to relieve the register pressure in loop nests where a large number of arrays is accessed. If the stiffness matrix can be evaluated by accessing a few (relatively ) basis functions or temporary values introduced by $licm$,

% why split useless in helmholtz/diffusion
% why vect vect+split useless in burgers

\begin{figure}[h]
\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmholtz kernel. Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh. CM stands for cost model.}
\label{fig:opvect-helmholtz-speedup}
\end{figure}

\begin{figure}[h]
\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Diffusion kernel. Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh. CM stands for cost model.}
\label{fig:opvect-diffusion-speedup}
\end{figure}

\begin{figure}[h]
\includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Burgers kernel. Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh. CM stands for cost model.}
\label{fig:opvect-burgers-speedup}
\end{figure}
